---
title: "Employee Absenteeism Analysis"
author: "Chau Anh Le"
date: "`r Sys.Date()`"

output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    theme: cosmo
runtime: shiny
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(readxl)
library(shiny)
library(DT)
library(flexdashboard)
library(ggplot2)
```

# 1. Introduction and Business Problem

- *Absenteeism_at_work* dataset.
- **Business Problem**: HR needs a reliable way to identify employee records that are more likely to be associated with high absenteeism so they can improve workforce planning and reduce productivity disruption.
- The variable of interest is **Absenteeism time in hours**. For predictive analysis, this variable is transformed into a binary outcome: **High absence** and **Low absence**.
- High absence is defined as 8 hours or more. This threshold is selected because the average absence time is around 6.9 hours, and 8 hours represents approximately one full working day.
- The model should be used as a decision-support tool for workforce planning, not as an automatic decision-maker for employees or job applicants.

# 2. Load and Inspect the Dataset

## Row {data-height="350"}

### **View first few rows**

```{r}
df <- read_excel("Absenteeism_at_work.xls")

head(df)
```

## Row {data-height="950"}

### **Data types**

```{r}
str(df)

```

### **Summary table**

```{r}
summary(df)
```

## 

-   The original dataset has 21 variables. The variable of interest is “Absenteeism time in hours”.
-   To support logistic regression, this numeric variable is transformed into a binary classification outcome: High absenteeism and Low absenteeism.

# 3. Data Preparation

## Row {data-height="600"}

### **Missing values**

```{r missing-values}
sum(is.na(df))
colSums(is.na(df))
```

-   No missing values

### **Duplicated rows**

```{r}
sum(duplicated(df))
dfc <- unique(df)
sum(duplicated(dfc))
dim(dfc)
```

-   34 duplicated rows were removed.

```{r}
# Remove invalid month value
dfc <- dfc[dfc$`Month of absence` != 0, ]
```

## Row {data-height="400"}

### **Classify "absenteeism hours" in to High and Low for further analysis**

```{r rename-variable}
names(dfc)[names(dfc) == "Absenteeism time in hours"] <- "absenteeism_hours"
```

```{r}
# Create binary target table for predictive analysis
dfc$high_absence <- ifelse(dfc$absenteeism_hours >= 8, "High", "Low")
# Convert to factor
dfc$high_absence <- factor(dfc$high_absence, levels = c("Low", "High"))
```

```{r}
# High and low absence distribution
table(dfc$high_absence)
prop.table(table(dfc$high_absence))
```

### 

```{r}
ggplot(as.data.frame(table(dfc$high_absence)), 
       aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(Var1, "\n", round(Freq / sum(Freq) * 100, 1), "%")),
            position = position_stack(vjust = 0.5),
            color = "white",
            size = 4) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Proportion of High and Low Absence",
    fill = "Absence Group"
  ) +
  theme_void()
```

## Row {data-height="650"}

### **Classify body mass index into 3 categories**

```{r categorise-bmi}
# Categorise bmi
dfc$bmi_category <- cut(
  dfc$`Body mass index`,
  breaks = c(-Inf, 18.5, 24.9, Inf),
  labels = c("Underweight", "Normal", "Overweight")
)
# Convert to factor
dfc$bmi_category <- as.factor(dfc$bmi_category)

table(dfc$bmi_category)
prop.table(table(dfc$bmi_category))
```

### **Employees' BMI distribution**

```{r bmi-dist}
ggplot(dfc, aes(x = bmi_category, fill = bmi_category)) + geom_bar() +
  labs(x = "BMI category", 
  y = "Number of employees",
  title ="Employee's BMI") +
  theme_bw()
```

## Row {data-height="650"}

### **Check outliers in variable of interest - Absenteeism in Hours**

```{r}
# Check outliers in variable of interest
boxplot.stats(dfc$absenteeism_hours)  # 44

```

### **Boxplot showing outliers in Absenteeism in Hours**

```{r}
# Boxplot before capping
boxplot(dfc$absenteeism_hours,
        main = "Boxplot for Absenteeism in Hours (Before)",
        xlab = "Absenteeism in Hours",
        ylab = "Hours")
```

### **Distribution of Absenteeism in Hours Before Capping**

```{r}
# Histogram before capping
ggplot(data = dfc, aes(x = absenteeism_hours)) + 
  geom_histogram() +
  labs(
    title = "Distribution of Absenteeism Hours Before Capping",
    x = "Absenteeism Hours",
    y = "Number of Records"
  ) +
  theme_bw()
```

## Row {data-height="650"}

### **Using IQR method to treat outliers**

```{r}
# IQR capping for absenteeism_hours
Q1 <- quantile(dfc$absenteeism_hours, 0.25)
Q3 <- quantile(dfc$absenteeism_hours, 0.75)
IQR_value <- IQR(dfc$absenteeism_hours)

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

lower_bound
upper_bound

# Identify outliers
outlier <- dfc$absenteeism_hours[dfc$absenteeism_hours < lower_bound | dfc$absenteeism_hours > upper_bound]
outlier

# Create capped version
dfc$absenteeism_hours_capped <- dfc$absenteeism_hours

dfc$absenteeism_hours_capped[dfc$absenteeism_hours < lower_bound] <- lower_bound
dfc$absenteeism_hours_capped[dfc$absenteeism_hours > upper_bound] <- upper_bound
```

### **Boxplot showing Absenteeism in Hours after Capping**

```{r}
# Boxplot after capping
boxplot(dfc$absenteeism_hours_capped,
        main = "Boxplot for Absenteeism Hours After Capping",
        xlab = "Absenteeism in Hours Capped",
        ylab = "Hours")
```

### **Distribution of Absenteeism in Hours After Capping**

```{r}
# Histogram after capping
ggplot(data = dfc, aes(x = absenteeism_hours_capped)) + 
  geom_histogram() +
  labs(
    title = "Distribution of Absenteeism Hours After Capping",
    x = "Capped Absenteeism Hours",
    y = "Number of Records"
  ) + 
  theme_bw()
```

```{r}
# Convert categorical variable
dfc$`Reason for absence` <- as.factor(dfc$`Reason for absence`)
dfc$`Month of absence` <- as.factor(dfc$`Month of absence`)
dfc$`Day of the week` <- as.factor(dfc$`Day of the week`)
dfc$Seasons <- as.factor(dfc$Seasons)
dfc$`Disciplinary failure` <- as.factor(dfc$`Disciplinary failure`)
dfc$Education <- as.factor(dfc$Education)
dfc$`Social drinker` <- as.factor(dfc$`Social drinker`)
dfc$`Social smoker` <- as.factor(dfc$`Social smoker`)

# Create clearer labels for Reason for absence
reason_labels <- c(
  "0" = "0 Unknown / not specified",
  "1" = "1 Infectious/parasitic diseases",
  "2" = "2 Neoplasms",
  "3" = "3 Blood/immune disorders",
  "4" = "4 Endocrine/metabolic diseases",
  "5" = "5 Mental/behavioural disorders",
  "6" = "6 Nervous system diseases",
  "7" = "7 Eye diseases",
  "8" = "8 Ear diseases",
  "9" = "9 Circulatory system diseases",
  "10" = "10 Respiratory diseases",
  "11" = "11 Digestive diseases",
  "12" = "12 Skin diseases",
  "13" = "13 Musculoskeletal diseases",
  "14" = "14 Genitourinary diseases",
  "15" = "15 Pregnancy/childbirth",
  "16" = "16 Perinatal conditions",
  "17" = "17 Congenital conditions",
  "18" = "18 Symptoms/signs not classified",
  "19" = "19 Injury/poisoning",
  "20" = "20 External causes",
  "21" = "21 Health service factors",
  "22" = "22 Patient follow-up",
  "23" = "23 Medical consultation",
  "24" = "24 Blood donation",
  "25" = "25 Laboratory examination",
  "26" = "26 Unjustified absence",
  "27" = "27 Physiotherapy",
  "28" = "28 Dental consultation"
)

dfc$reason_label <- reason_labels[as.character(dfc$`Reason for absence`)]
dfc$reason_label <- as.factor(dfc$reason_label)

# Create clearer labels for Education
education_labels <- c(
  "1" = "1 High school",
  "2" = "2 Graduate",
  "3" = "3 Postgraduate",
  "4" = "4 Master/doctor"
)

dfc$education_label <- education_labels[as.character(dfc$Education)]
dfc$education_label <- as.factor(dfc$education_label)

# Add labels to Seasons
dfc$season_label <- factor(
  dfc$Seasons,
  levels = c(1, 2, 3, 4),
  labels = c("Summer", "Autumn", "Winter", "Spring")
)

# Add labels to Day of the week
dfc$day_label <- factor(
  dfc$`Day of the week`,
  levels = c(2, 3, 4, 5, 6),
  labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
)
```

## Row {data-height="650"}

### **Summary statistics of the cleaned dataset**

```{r}
summary(dfc)
```

# 4. Descriptive Analytics

## Row {data-height=450}

```{r}
cat_choices <- c(
  "Reason for absence" = "reason_label",
  "Education" = "education_label",
  "Month of absence" = "Month of absence",
  "Day of the week" = "day_label",
  "Seasons" = "season_label",
  "Social drinker" = "Social drinker",
  "Social smoker" = "Social smoker",
  "BMI category" = "bmi_category"
)

selectInput(
  inputId = "selected_cat",
  label = "Choose a categorical variable to explore:",
  choices = cat_choices,
  selected = "reason_label"
)
```

```{r}
renderPlot({
  
  df_plot <- dfc
  df_plot$selected_variable <- df_plot[[input$selected_cat]]
  
  var_label <- names(cat_choices)[cat_choices == input$selected_cat]
  
  p <- ggplot(df_plot, aes(x = selected_variable, fill = high_absence)) +
    geom_bar(position = "fill") +
    labs(
      title = paste("Proportion of High and Low Absence by", var_label),
      x = var_label,
      y = "Proportion",
      fill = "Absence Group"
    ) +
    theme_bw()
  
  if (input$selected_cat == "reason_label") {
    p <- p + coord_flip()
  }
  
  p
})
```

```{r}
renderPlot({
  
  df_plot <- dfc
  df_plot$selected_variable <- df_plot[[input$selected_cat]]
  
  var_label <- names(cat_choices)[cat_choices == input$selected_cat]
  
  p <- ggplot(df_plot, aes(x = selected_variable, y = absenteeism_hours_capped)) +
    geom_boxplot() +
    labs(
      title = paste("Absenteeism Hours by", var_label),
      x = var_label,
      y = "Absenteeism Hours"
    ) +
    theme_bw()
  
  if (input$selected_cat == "reason_label") {
    p <- p + coord_flip()
  }
  
  p
})
```

## Row {data-height=250}

### **Select numeric variable**

```{r}
num_choices <- c(
  "Transportation expense" = "Transportation expense",
  "Distance from Residence to Work" = "Distance from Residence to Work",
  "Service time" = "Service time",
  "Age" = "Age",
  "Work load Average/day" = "Work load Average/day",
  "Hit target" = "Hit target",
  "Number of children" = "Son",
  "Number of pets" = "Pet",
  "Weight" = "Weight",
  "Height" = "Height",
  "Body mass index" = "Body mass index"
)

selectInput(
  inputId = "selected_num",
  label = "Choose a numeric variable to explore:",
  choices = num_choices,
  selected = "Age"
)
```

### **Correlation with absenteeism hours**

```{r}
renderPrint({
  
  selected_numeric <- dfc[[input$selected_num]]
  
  correlation_value <- cor(
    selected_numeric,
    dfc$absenteeism_hours,
    use = "complete.obs"
  )
  
  cat(
    "Correlation between",
    names(num_choices)[num_choices == input$selected_num],
    "and absenteeism hours:",
    round(correlation_value, 3)
  )
  
})
```

## Row {data-height=650}

### **Relationship with absenteeism hours**

```{r}
renderPlot({
  
  df_plot <- dfc
  df_plot$selected_numeric <- df_plot[[input$selected_num]]
  
  var_label <- names(num_choices)[num_choices == input$selected_num]
  
  ggplot(df_plot, aes(x = selected_numeric, y = absenteeism_hours_capped)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = paste("Relationship Between", var_label, "and Absenteeism Hours"),
      x = var_label,
      y = "Absenteeism Hours"
    ) +
    theme_bw()
  
})
```

### **Numeric variable by absence group**

```{r}
renderPlot({
  
  df_plot <- dfc
  df_plot$selected_numeric <- df_plot[[input$selected_num]]
  
  var_label <- names(num_choices)[num_choices == input$selected_num]
  
  ggplot(df_plot, aes(x = high_absence, y = selected_numeric, fill = high_absence)) +
    geom_boxplot() +
    labs(
      title = paste(var_label, "by Absence Group"),
      x = "Absence Group",
      y = var_label,
      fill = "Absence Group"
    ) +
    theme_bw()
  })
```

# 5. Predictive Analytics with Logistic Regression

## Row {data-height=620}

### **Prepare data for logistic regression**

```{r}
# Select variables for logistic regression model
model_data <- dfc[, c(
  "high_absence",
  "Education",
  "Son",
  "Age",
  "Transportation expense",
  "Work load Average/day",
  "Service time",
  "Month of absence",
  "Day of the week",
  "Seasons"
)]

# Make sure the target variable is a factor
model_data$high_absence <- factor(
  model_data$high_absence,
  levels = c("Low", "High")
)

# Make sure categorical predictors are factors
model_data$Education <- as.factor(model_data$Education)
model_data$`Month of absence` <- as.factor(model_data$`Month of absence`)
model_data$`Day of the week` <- as.factor(model_data$`Day of the week`)
model_data$Seasons <- as.factor(model_data$Seasons)

head(model_data)
str(model_data)
table(model_data$high_absence)
```

## Row {data-height=330}

### **Train and test split**

```{r}
set.seed(123)

train_index <- sample(
  1:nrow(model_data),
  size = 0.7 * nrow(model_data)
)

train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

nrow(train_data)
nrow(test_data)

table(train_data$high_absence)
table(test_data$high_absence)
```

## Row {data-height=3200}

### **Test different logistic regression models**

```{r}
# Model 1: simple employee-related model
model_1 <- glm(
  high_absence ~ Education + Son + Age + `Transportation expense`,
  data = train_data,
  family = binomial
)

# Model 2: full model with employee, work and time-related predictors
model_2 <- glm(
  high_absence ~ Education + Son + Age +
    `Transportation expense` +
    `Work load Average/day` +
    `Service time` +
    `Month of absence` +
    `Day of the week` +
    Seasons,
  data = train_data,
  family = binomial
)

# Model 3: remove weak numeric predictors
model_3 <- glm(
  high_absence ~ Education +
    `Transportation expense` +
    `Work load Average/day` +
    `Month of absence` +
    `Day of the week` +
    Seasons,
  data = train_data,
  family = binomial
)

# Model 4: final statistically focused model
model_4 <- glm(
  high_absence ~
    `Transportation expense` +
    `Work load Average/day` +
    `Month of absence` +
    `Day of the week` +
    Seasons,
  data = train_data,
  family = binomial
)

summary(model_1)
summary(model_2)
summary(model_3)
summary(model_4)
```

## Row {data-height=400}

### **Compare model performance**

```{r}
# Function to evaluate each model
evaluate_model <- function(model, test_data, cutoff = 0.50) {
  
  predicted_probability <- predict(
    model,
    newdata = test_data,
    type = "response"
  )
  
  predicted_class <- ifelse(
    predicted_probability >= cutoff,
    "High",
    "Low"
  )
  
  predicted_class <- factor(
    predicted_class,
    levels = c("Low", "High")
  )
  
  confusion_matrix <- table(
    Actual = test_data$high_absence,
    Predicted = predicted_class
  )
  
  accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
  
  false_positive <- confusion_matrix["Low", "High"]
  false_negative <- confusion_matrix["High", "Low"]
  
  sensitivity <- confusion_matrix["High", "High"] / sum(confusion_matrix["High", ])
  specificity <- confusion_matrix["Low", "Low"] / sum(confusion_matrix["Low", ])
  
  return(c(
    Accuracy = accuracy,
    False_Positive = false_positive,
    False_Negative = false_negative,
    Sensitivity = sensitivity,
    Specificity = specificity
  ))
}

# Function to count statistically significant coefficients
count_significant <- function(model) {
  
  p_values <- summary(model)$coefficients[, 4]
  p_values <- p_values[-1] # remove intercept
  
  significant_count <- sum(p_values < 0.05)
  
  return(significant_count)
}

model_eval <- rbind(
  evaluate_model(model_1, test_data),
  evaluate_model(model_2, test_data),
  evaluate_model(model_3, test_data),
  evaluate_model(model_4, test_data)
)

model_comparison <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4"),
  Description = c(
    "Simple employee-related model",
    "Full model",
    "Removed weak numeric predictors",
    "Final statistically focused model"
  ),
  AIC = c(
    AIC(model_1),
    AIC(model_2),
    AIC(model_3),
    AIC(model_4)
  ),
  Accuracy = model_eval[, "Accuracy"],
  False_Positive = model_eval[, "False_Positive"],
  False_Negative = model_eval[, "False_Negative"],
  Sensitivity = model_eval[, "Sensitivity"],
  Specificity = model_eval[, "Specificity"],
  Significant_Coefficients = c(
    count_significant(model_1),
    count_significant(model_2),
    count_significant(model_3),
    count_significant(model_4)
  )
)

model_comparison$AIC <- round(model_comparison$AIC, 2)
model_comparison$Accuracy <- round(model_comparison$Accuracy, 3)
model_comparison$Sensitivity <- round(model_comparison$Sensitivity, 3)
model_comparison$Specificity <- round(model_comparison$Specificity, 3)

datatable(model_comparison, rownames = FALSE)
```

## Row {data-height=1000}

### **Final model selection**

```{r}
# Select the final model
final_model <- model_4
```

Final selected model: **Model 4**

Reasons for selection:

- It has the lowest AIC among the tested models.
- It removes weaker predictors such as education, age, number of children and service time.
- It keeps statistically stronger predictors that are easier to explain to HR.
- It has similar accuracy to the fuller model, while being simpler.

Note:
The final model is not selected because every coefficient is significant. It is selected because it provides the best balance of model fit, accuracy, simplicity and business meaning.

```{r}
summary(final_model)
```

## Row {data-height=280}

### **Final model equation**

```{r}
coef_final <- coef(final_model)

# Start equation
equation_text <- paste0(
  "Probability of High Absence = 1 / (1 + exp(-(",
  round(coef_final["(Intercept)"], 4)
)

# Add each coefficient and variable into the equation
for (i in 2:length(coef_final)) {
  
  sign_symbol <- ifelse(coef_final[i] >= 0, " + ", " - ")
  
  equation_text <- paste0(
    equation_text,
    sign_symbol,
    format(abs(coef_final[i]), scientific = FALSE, digits = 5),
    " × ",
    names(coef_final)[i]
  )
}

# Close equation
equation_text <- paste0(equation_text, ")))")

cat(equation_text)
```
Business meaning:

- The model estimates the probability that an employee record belongs to the High Absence group.
- The final predictors are transportation expense, workload average per day, month of absence, day of the week and season.
- A higher predicted probability means the record is more likely to be classified as High Absence.


## Row {data-height=240}

### **Predict Probability of High absence for the Test Dataset**

```{r}
predicted_probability <- predict(
  final_model,
  newdata = test_data,
  type = "response"
)

head(predicted_probability)
summary(predicted_probability)
```

## Row {data-height=250}

### **Classify using cutoff 0.50**

```{r}
fixed_cutoff <- 0.50

predicted_class <- ifelse(
  predicted_probability >= fixed_cutoff,
  "High",
  "Low"
)

predicted_class <- factor(
  predicted_class,
  levels = c("Low", "High")
)

head(predicted_class)
table(predicted_class)
```

## Row {data-height=180}

### **Confusion matrix**

```{r}
confusion_matrix <- table(
  Actual = test_data$high_absence,
  Predicted = predicted_class
)

confusion_matrix
```

## Row {data-height=120}

### **Classification accuracy**

```{r}
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)

accuracy
```

## Row {data-height=120}

### **Interactive cutoff slider**

```{r}
sliderInput(
  inputId = "cutoff_value",
  label = "Choose probability cutoff for High Absence:",
  min = 0,
  max = 1,
  value = 0.5,
  step = 0.05
)
```

## Row {data-height=250}

### **Interactive confusion matrix**

```{r}
renderDT({
  
  predicted_probability <- predict(
    final_model,
    newdata = test_data,
    type = "response"
  )
  
  predicted_class <- ifelse(
    predicted_probability >= input$cutoff_value,
    "High",
    "Low"
  )
  
  predicted_class <- factor(
    predicted_class,
    levels = c("Low", "High")
  )
  
  confusion_matrix <- table(
    Actual = test_data$high_absence,
    Predicted = predicted_class
  )
  
  confusion_table <- as.data.frame.matrix(confusion_matrix)
  
  datatable(confusion_table, rownames = TRUE)
  
})
```

### **Accuracy and error rates**

```{r}
renderPrint({
  
  predicted_probability <- predict(
    final_model,
    newdata = test_data,
    type = "response"
  )
  
  predicted_class <- ifelse(
    predicted_probability >= input$cutoff_value,
    "High",
    "Low"
  )
  
  predicted_class <- factor(
    predicted_class,
    levels = c("Low", "High")
  )
  
  confusion_matrix <- table(
    Actual = test_data$high_absence,
    Predicted = predicted_class
  )
  
  accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
  
  class_1_error <- confusion_matrix["High", "Low"] / sum(confusion_matrix["High", ])
  class_0_error <- confusion_matrix["Low", "High"] / sum(confusion_matrix["Low", ])
  
  sensitivity <- confusion_matrix["High", "High"] / sum(confusion_matrix["High", ])
  specificity <- confusion_matrix["Low", "Low"] / sum(confusion_matrix["Low", ])
  
  cat("Cutoff value:", input$cutoff_value, "\n")
  cat("Accuracy:", round(accuracy, 3), "\n")
  cat("Class 1 error rate - actual High predicted Low:", round(class_1_error, 3), "\n")
  cat("Class 0 error rate - actual Low predicted High:", round(class_0_error, 3), "\n")
  cat("Sensitivity - correctly predicted High:", round(sensitivity, 3), "\n")
  cat("Specificity - correctly predicted Low:", round(specificity, 3), "\n")
  
})
```

# 6. HR Prediction Tool and Recommendations

## Row {data-height=350}

### **Enter employee/applicant information**

```{r}
numericInput(
  inputId = "new_transport",
  label = "Transportation expense:",
  value = 200,
  min = 0
)

numericInput(
  inputId = "new_workload",
  label = "Workload average per day:",
  value = 270000,
  min = 0
)
```

### **Select time-related information**

```{r}
selectInput(
  inputId = "new_month",
  label = "Month of absence:",
  choices = c(
    "January" = "1",
    "February" = "2",
    "March" = "3",
    "April" = "4",
    "May" = "5",
    "June" = "6",
    "July" = "7",
    "August" = "8",
    "September" = "9",
    "October" = "10",
    "November" = "11",
    "December" = "12"
  ),
  selected = "7"
)

selectInput(
  inputId = "new_day",
  label = "Day of the week:",
  choices = c(
    "Monday" = "2",
    "Tuesday" = "3",
    "Wednesday" = "4",
    "Thursday" = "5",
    "Friday" = "6"
  ),
  selected = "2"
)

selectInput(
  inputId = "new_season",
  label = "Season:",
  choices = c(
    "Summer" = "1",
    "Autumn" = "2",
    "Winter" = "3",
    "Spring" = "4"
  ),
  selected = "1"
)
```

### **Choose cutoff threshold**

```{r}
sliderInput(
  inputId = "new_cutoff",
  label = "Choose cutoff threshold for High Absence:",
  min = 0,
  max = 1,
  value = 0.5,
  step = 0.05
)
```

## Row {data-height=300}

### **Prediction result**

```{r}
renderPrint({
  
  # Create new record using the same column names as the model
  new_record <- data.frame(
    `Transportation expense` = input$new_transport,
    `Work load Average/day` = input$new_workload,
    `Month of absence` = factor(
      input$new_month,
      levels = levels(train_data$`Month of absence`)
    ),
    `Day of the week` = factor(
      input$new_day,
      levels = levels(train_data$`Day of the week`)
    ),
    Seasons = factor(
      input$new_season,
      levels = levels(train_data$Seasons)
    ),
    check.names = FALSE
  )
  
  # Predict probability of High Absence
  predicted_probability_new <- predict(
    final_model,
    newdata = new_record,
    type = "response"
  )
  
  # Classify using selected cutoff
  predicted_class_new <- ifelse(
    predicted_probability_new >= input$new_cutoff,
    "High Absence",
    "Low Absence"
  )
  
  cat("Predicted probability of High Absence:", 
      round(predicted_probability_new, 3), "\n")
  
  cat("Selected cutoff threshold:", 
      input$new_cutoff, "\n")
  
  cat("Predicted classification:", 
      predicted_class_new, "\n")
  
})
```

### **HR recommendation based on prediction**

```{r}
renderPrint({
  
  new_record <- data.frame(
    `Transportation expense` = input$new_transport,
    `Work load Average/day` = input$new_workload,
    `Month of absence` = factor(
      input$new_month,
      levels = levels(train_data$`Month of absence`)
    ),
    `Day of the week` = factor(
      input$new_day,
      levels = levels(train_data$`Day of the week`)
    ),
    Seasons = factor(
      input$new_season,
      levels = levels(train_data$Seasons)
    ),
    check.names = FALSE
  )
  
  predicted_probability_new <- predict(
    final_model,
    newdata = new_record,
    type = "response"
  )
  
  predicted_class_new <- ifelse(
    predicted_probability_new >= input$new_cutoff,
    "High Absence",
    "Low Absence"
  )
  
  if (predicted_class_new == "High Absence") {
    
    cat("Recommendation:\n")
    cat("This record is predicted as High Absence.\n")
    cat("HR should consider workload backup, flexible staffing, and early support planning.\n")
    cat("This prediction should support planning only and should not be used as an automatic hiring or employee decision.\n")
    
  } else {
    
    cat("Recommendation:\n")
    cat("This record is predicted as Low Absence.\n")
    cat("HR can proceed with normal workforce planning.\n")
    cat("However, the result should still be reviewed together with human judgement and other workplace information.\n")
    
  }
  
})
```

## Row {data-height=450}

### **Overall recommendations to HR**

1. **Use the model as a planning tool, not an automatic decision-maker**  
The model should support workforce planning and absence risk monitoring. It should not be used to automatically reject applicants or penalise employees.

2. **Adjust the cutoff based on business priority**  
A lower cutoff identifies more possible High Absence cases, reducing missed high-risk records but increasing false positives. A higher cutoff is stricter and reduces false positives but may miss more actual High Absence cases.

3. **Prepare staffing buffers for high-risk periods**  
Month and season are important predictors, so HR should prepare extra workload coverage during higher-risk months or seasons.

4. **Monitor workload-related risk**  
Workload average per day is included in the final model, suggesting workload patterns may help explain High Absence probability.

5. **Use human judgement with model output**  
The model achieved moderate accuracy, so predictions should be reviewed with context, fairness, and HR judgement before action is taken.
